Some thoughts on Zermelo's iteration process in the context of the hyper2 package

### 2023-10-30 inst/T20.Rmd
label all chunks, cache the slow ones

```{r indstrength}
attach(lapply(T20_table[, ], as.vector))  # vector access to team1, etc
T20 <- hyper2()
for(i in seq_along(team1)){
   T20[match_winner[i]] %<>% inc
   T20[c(team1[i],team2[i])] %<>% dec

@@

```{r nullallteams}
a1 <- maxp(T20,give=TRUE)$value        # likelihood at the evaluate
a2 <- loglik(indep(equalp(T20)),T20)  # likelihood at p1=p2=...=p14
a1-a2
```

@@

```{r tossdiff}
T20_toss  <- hyper2(pnames=c('toss',levels(T20_table$team1)))
for(i in seq_along(team1)){
      players <- c(team1[i],team2[i])
      if(toss_winner[i] == match_winner[i]){ # win toss, win match
         T20_toss[c('toss',match_winner[i])] %<>% inc
      } else { # win toss, lose match
         T20_toss[match_winner[i]] %<>% inc
      }			 
   T20_toss[c('toss',players)] %<>% dec
   T20_toss
}			  
```
and again we can maximize the likelihood:

>>>>

```{r indstrength,cache=TRUE}
attach(lapply(T20_table[, ], as.vector))  # vector access to team1, etc
T20 <- hyper2()
for(i in seq_along(team1)){
   T20[match_winner[i]] %<>% inc
   T20[c(team1[i],team2[i])] %<>% dec

@@

```{r nullallteams,cache=TRUE}
a1 <- maxp(T20,give=TRUE)$value        # likelihood at the evaluate
a2 <- loglik(indep(equalp(T20)),T20)  # likelihood at p1=p2=...=p14
a1-a2
```

@@

```{r tossdiff,cache=TRUE}
T20_toss  <- hyper2(pnames=c('toss',levels(T20_table$team1)))
for(i in seq_along(team1)){
      players <- c(team1[i],team2[i])
      if(toss_winner[i] == match_winner[i]){ # win toss, win match
         T20_toss[c('toss',match_winner[i])] %<>% inc
      } else { # win toss, lose match
         T20_toss[match_winner[i]] %<>% inc
      }			 
   T20_toss[c('toss',players)] %<>% dec
   T20_toss
}			  
```
and again we can maximize the likelihood:

### 2023-10-30 inst/formula1_points.Rmd
minor typos
label all chunks, cache the slow ones

```{r}
f <- function(x){
	noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
	x <- as.character(x)
	x[x %in% noscore] <-  "0"
	x <- as.numeric(x)
	wanted <- x != 0
	xw <- x[wanted]
	xw[order(xw,decreasing=FALSE)] <- seq_along(xw)
	x[wanted] <- xw
	x
}
o  <- apply(f2017[,seq_len(20)],2,f)
rownames(o) <- rownames(f2017)
(f2017 <- o)
```

@@

```{r}
points <- c(25,18,15,12,10,8,6,4,2,1,0,0,0,0,0,0,0,0,0,0)
stopifnot(all(diff(points)<=0))
points <- points/points[1]
jj <- f2017
jj[jj==0] <- length(points)
jj[] <- points[jj]
jj
real_points <- rowSums(jj)
real_points
sort(real_points,decreasing=TRUE)
names(sort(real_points,decreasing=TRUE))
```

@@

```{r}
random_table <- rrank(n=ncol(f2017), p=mL2017)
rownames(random_table) <- colnames(f2017)
random_table
```

@@

```{r}
maxp(rank_likelihood(random_table)) # should be close to mL2017
```

```{r}
a <- random_table
class(a) <- "matrix"
a[] <- points[a]
a
p <- colSums(a)
p
rando <- order(p,decreasing=TRUE)
rando
```

@@

```{r}
jj <- sort(p,decreasing=TRUE)
jj
names(jj)
```

@@

>>>>

```{r processtogetranks}
f <- function(x){
	noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
	x <- as.character(x)
	x[x %in% noscore] <-  "0"
	x <- as.numeric(x)
	wanted <- x != 0
	xw <- x[wanted]
	xw[order(xw,decreasing=FALSE)] <- seq_along(xw)
	x[wanted] <- xw
	x
}
o  <- apply(f2017[,seq_len(20)],2,f)
rownames(o) <- rownames(f2017)
(f2017 <- o)
```

@@

```{r calcthepoints}
points <- c(25,18,15,12,10,8,6,4,2,1,0,0,0,0,0,0,0,0,0,0)
stopifnot(all(diff(points)<=0))
points <- points/points[1]
jj <- f2017
jj[jj==0] <- length(points)
jj[] <- points[jj]
jj
real_points <- rowSums(jj)
real_points
sort(real_points,decreasing=TRUE)
names(sort(real_points,decreasing=TRUE))
```

@@

```{r userandomsamp}
random_table <- rrank(n=ncol(f2017), p=mL2017)
rownames(random_table) <- colnames(f2017)
random_table
```

@@

```{r howclosetoreality,cache=TRUE}
maxp(rank_likelihood(random_table)) # should be close to mL2017
```

```{r checkagain}
a <- random_table
class(a) <- "matrix"
a[] <- points[a]
a
p <- colSums(a)
p
rando <- order(p,decreasing=TRUE)
rando
```

@@

```{r thuswesee}
jj <- sort(p,decreasing=TRUE)
jj
names(jj)
```

@@

# Now account for pole position

Now do the same thing again but include pole position.

### 2023-10-30 inst/handover.Rmd
use sticker in Rmd
label all chunks, cache the slow ones

```{r}
handover <- hyper2()
handover["t11"] %<>% inc(handover_table[1,1])
handover["t10"] %<>% inc(handover_table[1,2])
handover["t01"] %<>% inc(handover_table[2,1])
handover["t00"] %<>% inc(handover_table[2,2])

@@

```{r}
samep.test(handover,c("t01","t10"))
```

@@

```{r}
probability(handover,disallowed=function(p){p[2]<p[3]},tol=0.05)
```

>>>>

```{r setuphandover,cache=TRUE}
handover <- hyper2()
handover["t11"] %<>% inc(handover_table[1,1])
handover["t10"] %<>% inc(handover_table[1,2])
handover["t01"] %<>% inc(handover_table[2,1])
handover["t00"] %<>% inc(handover_table[2,2])

@@

```{r testtheta01,cache=TRUE}
samep.test(handover,c("t01","t10"))
```

@@

```{r probofhandover}
probability(handover,disallowed=function(p){p[2]<p[3]},tol=0.05)
```

### 2023-10-30 inst/loser.Rmd
slight rewriting
label all chunks, cache the slow ones

Here we consider a situation in which we make only a rather
uninformative observation: namely, the identity of the _loser_ (which
is here taken to be me, player `a`).  We have six competitors
`a,b,c,d,e,f` with nonnegative strengths $p_1,p_2,p_3,p_4,p_5,p_6,
\sum p_i=1$.  Competitor `a` has strength $p_1$, and loses.  We are
interested in how informative this observation is with regard to
various assertions about the strength of `a`.  We have three
hypotheses:

@@

```{r,label=useggrl}
n <- 6 # number of competitors
H <- hyper2(pnames=letters[1:6])
L <- ggrl(H,letters[2:n],letters[1]) 
length(L)
L[[1]]
L[[2]]
```

>>>>

(see also `inst/notthewinner.Rmd`).
Here we consider a situation in which we make only a rather
uninformative observation: namely, the identity of the _loser_ (which
is here taken to be me, player `a`).  We have six competitors
`a,b,c,d,e,f` with nonnegative strengths $p_1,p_2,p_3,p_4,p_5,p_6,
\sum p_i=1$.  Competitor `a` has strength $p_1$, and loses.  We are
interested in how informative this observation is with regard to
various assertions about the strength of `a`.  We have three
hypotheses:

@@

```{r,label=useggrl,cache=TRUE}
n <- 6 # number of competitors
H <- hyper2(pnames=letters[1:6])
L <- ggrl(H,letters[2:n],letters[1]) 
length(L)
L[[1]]
L[[2]]
```

### 2023-10-30 inst/order.Rmd
label all chunks, cache the slow ones

```{r}
evaluate1 <- maxp(H)
(maxlike1 <- loglik(indep(evaluate1),H))
dotchart(evaluate1,pch=16)
```

@@

```{r}
f <- function(p,n){c(p,rep((1-p)/n,n-1))}
sss <- f(0.99,9)  # sensible start point, close (?) to the evaluate
evaluate2 <- maxp(H,startp=sss)
maxlike2 <- loglik(indep(evaluate2),H)
```

@@

```{r}
dotchart(evaluate2,pch=16)
par(pty='s')
plot(evaluate1,evaluate2,asp=1,log='xy')
abline(0,1)
```

@@

```{r}
maxlike2-maxlike1
```

@@

```{r}
vsssp <- function(x,n){
  jj <- 1/x^seq_len(n)
  indep(jj/sum(jj))
}
```

@@

```{r}
evaluate3 <- maxp(H,vsssp(2,10))
maxlike3 <- loglik(indep(evaluate3),H)
dotchart(evaluate3,pch=16)
```

@@

```{r}
maxlike3-maxlike2
```

>>>>

```{r,label=maxwithevalp,cache=TRUE}
evaluate1 <- maxp(H)
(maxlike1 <- loglik(indep(evaluate1),H))
dotchart(evaluate1,pch=16)
```

@@

```{r bettereval,cache=TRUE}
f <- function(p,n){c(p,rep((1-p)/n,n-1))}
sss <- f(0.99,9)  # sensible start point, close (?) to the evaluate
evaluate2 <- maxp(H,startp=sss)
maxlike2 <- loglik(indep(evaluate2),H)
```

@@

```{r showmaxlike}
dotchart(evaluate2,pch=16)
par(pty='s')
plot(evaluate1,evaluate2,asp=1,log='xy')
abline(0,1)
```

@@

```{r diffbetweenlikelihoods}
maxlike2-maxlike1
```

@@

```{r mimiclimit}
vsssp <- function(x,n){
  jj <- 1/x^seq_len(n)
  indep(jj/sum(jj))
}
```

@@

```{r evaluatetwice,cache=TRUE}
evaluate3 <- maxp(H,vsssp(2,10))
maxlike3 <- loglik(indep(evaluate3),H)
dotchart(evaluate3,pch=16)
```

@@

```{r veryslightlybetter}
maxlike3-maxlike2
```

### 2023-10-30 inst/pentathlon.Rmd
label all chunks, cache the slow ones

```{r}
pentathlon <- ordertable2supp(pentathlon_table)
pentathlon_maxp <- maxp(pentathlon)
pentathlon_maxp
pie(pentathlon_maxp)
```

@@

```{r}
equalp.test(pentathlon)
```

>>>>

```{r showpiechart,cache=TRUE}
pentathlon <- ordertable2supp(pentathlon_table)
pentathlon_maxp <- maxp(pentathlon)
pentathlon_maxp
pie(pentathlon_maxp)
```

@@

```{r testequal,cache=TRUE}
equalp.test(pentathlon)
```

### 2023-10-30 inst/zermelo.Rmd
label all chunks, cache the slow ones

```{r}
n <- 10
p <- zipf(n)
names(p) <- letters[seq_len(n)]
p
```

@@

```{r}
M <- matrix(0,n,n)
rownames(M) <- names(p)
colnames(M) <- names(p)
index <- which(upper.tri(M),arr.ind=TRUE)
for(o in seq_len(nrow(index))){
  i <- index[o,1]
  j <- index[o,2]
  prob <- p[i]/(p[i] + p[j]) # Bradley-Terry
  nobs <- 80 
  i_wins <- rbinom(1,nobs,prob)
  j_wins <- nobs - i_wins
  M[i,j] <- i_wins
  M[j,i] <- j_wins
}
M
```

@@

```{r}
(pZ <- zermelo(M))
```

@@

```{r,cache=TRUE}
(H <- pairwise(M))
(pH <- maxp(H))
```

@@

```{r}
par(pty='s')
plot(pZ,pH,asp=1,log='xy')
abline(0,1)
```

```{r}
plot(log(pZ),log(abs(pH-pZ)/pZ))
abline(h=0)
```

```{r}
loglik(pZ,H)-loglik(pH,H)
```

@@

```{r}
par(pty='s')
plot(zipf(n),log(abs(zipf(n)-pZ)/zipf(n)),pch=16,col='black',log='x')
```

@@

```{r}
o <- zermelo(M,give=TRUE)
head(o)
l <- apply(o,1,loglik,H)
plot(log(abs(l-max(l))))
```

>>>>

```{r maketencompetitors}
n <- 10
p <- zipf(n)
names(p) <- letters[seq_len(n)]
p
```

@@

```{r makeMwithpairwise}
M <- matrix(0,n,n)
rownames(M) <- names(p)
colnames(M) <- names(p)
index <- which(upper.tri(M),arr.ind=TRUE)
for(o in seq_len(nrow(index))){
  i <- index[o,1]
  j <- index[o,2]
  prob <- p[i]/(p[i] + p[j]) # Bradley-Terry
  nobs <- 80 
  i_wins <- rbinom(1,nobs,prob)
  j_wins <- nobs - i_wins
  M[i,j] <- i_wins
  M[j,i] <- j_wins
}
M
```

@@

```{r usezerm}
(pZ <- zermelo(M))
```

@@

```{r,label=usehyper2,cache=TRUE}
(H <- pairwise(M))
(pH <- maxp(H))
```

@@

```{r comparewithscattergraph}
par(pty='s')
plot(pZ,pH,asp=1,log='xy')
abline(0,1)
```

```{r nowplotlog}
plot(log(pZ),log(abs(pH-pZ)/pZ))
abline(h=0)
```

```{r calcloglik}
loglik(pZ,H)-loglik(pH,H)
```

@@

```{r arebothclose}
par(pty='s')
plot(zipf(n),log(abs(zipf(n)-pZ)/zipf(n)),pch=16,col='black',log='x')
```

@@

```{r funwithzermelo}
o <- zermelo(M,give=TRUE)
head(o)
l <- apply(o,1,loglik,H)
plot(log(abs(l-max(l))))
```
